The Limits of Loyalty: Postmodernism, Polarization, and Military Professionalism  (Prospectus)

### 2020-09-23 code/prospectusv1.Rmd
commit

# The Scholarly Literature and Why This Question Matters

The question of how political polarization affects changes in the concept of military professionalism spans several related literatures, most notable of which are American Politics and International Relations. Broadly speaking, this project is focused on how one aspect of domestic politics, political polarization, impacts a specific but important aspect of international relations, civil-military relations, which itself spans a wide and diverse literature. Because I intend to focus my work on the military institution, my dissertation will also rely on literature from the fields of sociology, mainly regarding the development of organizations and professions, and ethics, to include military ethics.

Within the American Politics literature, there is broad recognition that political polarization levels in contemporary America are higher now than at any point since the end of the American Civil War [@hare_polarization_2014], but there are multiple explanations for what has caused this trend, ranging from a growing divide over matters pertaining to race and religion [@abramowitz_great_2018], the size and role of government [@sides_two_2015], and disagreements over specific policy issues [@mann_its_2012]. Despite the herculean task of disentangling the roots of the causes of polarization, virtually all scholars agree that political polarization is a genuine problem and should spark concern.

One of these major concerns is the impact of polarized domestic politics on American foreign policy in a broad sense, and specifically, on America's ability to lead and impact the world. Kenneth Schultz, for instance, underscored four challenges that have emerged as a result of America's polarized domestic politics, including the ability to obtain bipartisan support for "risky undertakings," the ability to learn the "lessons of foreign policy failure," the inability to "make long term commitments to allies and adversaries" as a result of "dramatic policy swings from one administration to another," and the "vulnerability" of America's electoral system to foreign third party adversaries [@schultz_perils_2017]. Because the military is a central actor in shaping America's foreign policy, obtaining a better understanding of how political polarization shapes various military institutions specifically will yield findings that may bear on helping policymakers and academics help solve each of the problems Schultz identifies.

Thankfully, a small but growing number of civil-military relations scholars have looked at aspects of this question in recent years. But the focus of these works has been, I believe, not on the impact of polarization on the military itself but on other important but peripheral components of the military's relationship with American society. One recent dissertation completed by Michael Robinson at Stanford University in 2018, for instance, found that partisan affiliation significantly affects public credibility and trust in the military perhaps resulting in or as a consequence of media "echo chambers" which may incentivize military leaders to offer opinions partisan opinions without fear of sanction [@robinson_danger_2018]. Robinson's work builds off of previous work by Heidi Urben in which she found that active military officers believe less strongly that retired military officers should remain apolitical or non-partisan in retirement [@urben_party_2013], a norm that has existed for decades or more. Together, Robinson and Urben join other scholars in providing a wealth of information regarding public opinion and the military, and in particular, how aspects of political polarization shape partisan differences regarding topics such as the use of force [@golby_jim_elite_2017] and military elite endorsement of political candidates [@golby_military_2012].

In their own right, these scholarly developments are interesting and important, but another debate in the civil-military relations field has recently emerged, or to be more correct, reawakened. At the center of this debate is the degree to which the concept of military professionalism should be altered if not discarded altogether [@brooks_paradoxes_2020]. The debate is not new because in actuality, it began as soon as Samuel Huntington published his seminal work in 1957, _The Soldier and the State_. While I do not lay out the full synopsis of the book nor the the various critiques here explicitly, the central question in this debate is whether the concept of military professionalism, which Huntington espoused as essential to achieving military security during the Cold War and as the US military has largely inculcated among its officer corps for the last couple of generations, separates the military too much from civilians, leading to gaps in the development and execution of strategic outcomes [@rapp_civil-military_2015; @dubik_just_2016; @brooks_paradoxes_2020]. These criticisms largely point to the American failures in Iraq and Afghanistan to produce and implement strategy and achieve a better and lasting peace - criticisms which this author, for what it is worth, shares. But I contend that something is missing in this debate, namely, a thorough exploration and theoretical outworking of the impacts of political polarization on the military institution.

Why does this matter? I contend that the impacts of contemporary political polarization on civil-military relations and on the military institution specifically matters and should be studied more thoroughly because, as I intend to show, portions of the military institution, though not all, have started to argue and disagree with civilian leaders in ways that are substantively different than previous civil-military feuds. To be sure, military and civilian relationships are in tension by their very nature in that the leaders of a democracy allow a most undemocratic institution, the military, to protect the polity [@feaver_civil-military_1996]. And as American history proves, there have been countless instances in which civilian and military leaders disagreed. Many of these rise to the level of common knowledge: Lincoln fired McClellan because he lost confidence in him as a wartime Commander; Truman sacked MacArthur because the latter was insubordinate regarding the future of war in Asia; and Obama and McChrystal clashed about troop levels in Afghanistan.

Beyond these episodes in which President and wartime commander locked horns, consider several other instances of "civilians" and "military" clashing. One episode is certainly the way in which the Vietnam war escalated in ways that, at least according to H.R. McMaster, military generals at the time knew were destined to fail [@mcmaster_dereliction_1998]. A more recent example, and one much studied and commented upon by civil-military relations scholars, is the so-called "Revolt of the Generals" that occurred following the early years of the invasion of Iraq, which involved several recently-retired generals and admirals sharply criticizing Secretary of Defense Rumsfeld's leadership dynamics [@cook_revolt_2008]. The point here is that civil-military conflict is anything but novel, but that when it occurs, it is usually, for understandable reasons, involving a clear national security issue, such as a war, combat operations, or items of a related sort, such as military budgets, weapons programs, etc.

While these types of disagreements are important and will certainly feature in my intended dissertation (I hope!), my motivation in choosing my intended topic is because we are now seeing portions of the military institution (particular retired military officers) openly disagree with civilian leaders on topics that I believe are, even by the widest of academic definitions, way outside of appropriate military expertise. In other words, there is a world of difference between retired generals criticizing a President for conduct of a war, but entirely another when generals criticize (or applaud) the President for his domestic policies that have little if anyting to do with national security. I contend that the latter is taking place, and that the reasons for such developments, as well as their implications, deserve to be further explored. I further believe that while such developments are catching the attention of civil-military relations scholars, most of the work done on this topic is impressionistic [@golby_jim_keep_2020] as opposed to systematic. In other words, civil-military relations scholars have identified that polarized partisan politics serve as a risk for the military, but few have explored in a systematic way what these risks are, how serious they are, the mechanisms through which they work, and what the implications are for the future of the military. I hope to fill this gap. In doing so, I hope to also contribute to the reawakened debate regarding military professionalism and its role vis-a-vis the American military institution.

## Polarization and Military Professionalism

>>>>

# Framing Within the Scholarly Literature

## Political Polarization

The question of if and how political polarization drives changes in the professional identity of the military spans several related literatures, most notable of which are American Politics and International Relations. Broadly speaking, this project is focused on how one aspect of domestic politics, political polarization, impacts a specific but important aspect of international relations, civil-military relations, which itself spans a wide and diverse literature. Because I intend to focus my work on the military as an institution, my dissertation will also rely on literature from the fields of sociology, organizational psychology, and ethics. I will frame my research question within each of these fields briefly before addressing why this question matters and should be studied in depth.

Within the American Politics literature, there is broad recognition that political polarization levels in contemporary America are higher now than at any point since the end of the American Civil War [@hare_polarization_2014], but there are multiple explanations for what has caused this trend, ranging from a growing divide over matters pertaining to race and religion [@abramowitz_great_2018], the size and role of government [@sides_two_2015], and disagreements over specific policy issues [@mann_its_2012]. Despite the herculean task of disentangling the roots of the causes of polarization, scholars agree that political polarization is a problem of sorts.

One of the major concerns regarding political polarization is its on impact on American foreign policy in a broad sense, and specifically, on America's ability to lead and impact the world. Kenneth Schultz, for instance, underscored four challenges that have emerged as a result of America's polarized domestic politics, including the ability to obtain bipartisan support for "risky undertakings," the ability to learn the "lessons of foreign policy failure," the inability to "make long term commitments to allies and adversaries" as a result of "dramatic policy swings from one administration to another," and the "vulnerability" of America's electoral system to foreign third party adversaries [@schultz_perils_2017]. Because the military is a central actor in shaping America's foreign policy, obtaining a better understanding of how political polarization shapes various military institutions specifically will yield findings that may bear on helping policymakers and academics help solve each of the problems Schultz identifies.

## Civil Military Relations

Thankfully, a small but growing number of civil-military relations scholars have looked at aspects of this question in recent years. But the focus of these works has been, I believe, not on the impact of polarization on the military itself but on other important but peripheral components of the military's relationship with American society. These scholars have made important findings that speak to the state of current American civil-military relations. One recent dissertation completed by Michael Robinson at Stanford University in 2018, for instance, found that partisan affiliation significantly affects public credibility and trust in the military perhaps resulting in or as a consequence of media "echo chambers" which may incentivize military leaders to offer opinions partisan opinions without fear of sanction [@robinson_danger_2018]. Robinson's work builds off of previous work by Heidi Urben in which she found that active military officers believe less strongly that retired military officers should remain apolitical or non-partisan in retirement [@urben_party_2013], a norm that has existed for decades or more. Together, Robinson and Urben join other scholars in providing a wealth of information regarding public opinion and the military, and in particular, how aspects of political polarization shape partisan differences regarding topics such as the use of force [@golby_jim_elite_2017] and military elite endorsement of political candidates [@golby_military_2012].

In their own right, these scholarly developments are interesting and important, but another debate in the civil-military relations field has recently emerged, or to be more correct, reawakened. At the center of this debate is the degree to which the concept of _military professionalism_ should be altered, if not discarded altogether [@brooks_paradoxes_2020], by the collective civilian-military leadership, in part due to concerns about the tense domestic political environment that exists in contemporary America. The central question in this debate is whether the concept of _military professionalism_ actually accomplishes its intended goals: to simultaneously achieve military security from external threats, and to maintain civilian control of the military -- an argument developed in Samuel Huntington's 1957 work, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_.

In this seminal work, Huntington asserts that two distinct forces would always be in tension with one another in a liberal democracy such as the United States. The first was a "functional imperative," which Huntington said drove the requirement that a military be strong enough and capable of defeating the military threats it faced, and the second was a "societal imperative," which Huntington claims constrains the military in that the military can never lose certain aspects such that the military no longer represents the liberal democracy it defends [@huntington_soldier_1957, 1-3]. Writing in the context of the seemingly long and dangerous Cold War, Huntington's point is that America needs a strong military that can win on the battlefield, but cautions that such a military cannot fundamentally alienate the character of American liberal society.

Huntington asserts that the key to resolving the tension generated by these two opposing forces - to win on the battlefield and to maintain itself as a liberal democracy - is to adopt a scheme of "objective civilian control" whereby political leaders maximize the professional autonomy of the military by establishing the ends and goals of policy, but allowing military leaders to develop and implement the military plans to accomplish these goals [@huntington_soldier_1957, 80-85]. The creation and maintenance of largely separate operating spheres between the political leaders of a democracy and the military as an institution would ensure that the military would win on the battlefield, but also stay out of politics and thus, untainted by partisan interests and concerns.

Numerous critiques surrounding the Huntingtonian ideal of _military professionalism_ have been levied over the years. Most of these critiques, in one way or another, charge that a strict theoretical and practical separation between military and political operating spheres is fundamentally flawed. Many, if not all, of these critiques cite Clausewitz's famous dictum that war is a "mere continuation of policy by other means" [@clausewitz_war_1976, 69, 605], charging that war by its very nature requires the highest integration - not separation - of military and political activities. In essence, these critiques argue that the strict separation of the responsibilities between the military and civilians leads to poor strategic outcomes [@rapp_civil-military_2015; @brooks_paradoxes_2020].  Moreover, these criticisms can and do point largely to American failures in Iraq and Afghanistan to secure, at least as of yet, a more-lasting peace - failures that are ultimately, according to one scholar and former general, moral in nature, and shared between both civilians and the military [@dubik_just_2016].

## The Concept and Norm of Organizational Loyalty

An important value among militaries, especially professional militaries, is loyalty. Indeed, the US Army's first "Army Value" is "loyalty," followed by six other traits that in totality, spell out the acronym, l-d-r-s-h-i-p, or "leadership": "integrity," "duty," "respect," selfless service," "honor," integrity," and "personal courage" [need cite here]. The US Army states in its doctrinal manual on Army Leadership and the Profession, ADP 6-22, that the Army Values "embody the practical application of the Army Ethic. They encompass the enduring moral principles, beliefs, and laws that guide Army professionals in accomplishing the mission as well as their conduct in all aspects of life" [need cite here for ADP 6-22, p. 1-12]. The manual defines loyalty as, "loyalty: bear true faith and allegiance to the Constitution of the United States, the Army, your unit and other Soldiers" [need cite here for ADP 6-22.]

# Where My Puzzle Fits and Why It Matters

One area of work that remains under-developed, both theoretically and empirically, is an exploration of the relationship between polarizing domestic politics and the professional identity of the military, both as an American institution, but also as a profession. Exploring this relationship will, I believe, help scholars in the field of civil-military relations and international relations more broadly better understand if and how the military, as a unique body charged with the defense of the nation, changes in fundamental ways as a result of a polarized political climate. On one hand, one would expect that it would, in the sense that the military, at least in America, is comprised of people from all walks of life. On the other hand, one might expect that the unique purpose and "professional" nature of the military might insulate it from some of the impacts of polarized politics. I am especially interested, to be more specific, in if and how political polarization impacts the concept of organizational loyalty in the military, particularly because one of the hallmarks of the military's professional nature is that it ought to be apolitical and non-partisan. Yet in an environment where the ability to be apolitical and non-partisan is becoming more and more difficult, due to rising political polarization, it is important that scholars and policymakers understand the mechanisms through which the military is being shaped, if there any, resulting in changes to its professional character.

To be clear, scholars have long studied how domestic politics impact the military. To be sure, military and civilian relationships are in tension by their very nature in that the leaders of a democracy allow a most undemocratic institution, the military, to protect the polity [@feaver_civil-military_1996]. And as American history proves, there have been countless instances in which civilian and military leaders disagreed. Many of these rise to the level of common knowledge: Lincoln fired McClellan because he lost confidence in him as a wartime Commander; Truman sacked MacArthur because the latter was insubordinate regarding the future of war in Asia; and Obama and McChrystal clashed about troop levels in Afghanistan. Beyond these episodes in which President and wartime commander locked horns, consider several other instances of "civilians" and "military" clashing. One episode is certainly the way in which the Vietnam war escalated in ways that, at least according to H.R. McMaster, military generals at the time knew were destined to fail [@mcmaster_dereliction_1998]. A more recent example, and one much studied and commented upon by civil-military relations scholars, is the so-called "Revolt of the Generals" that occurred following the early years of the invasion of Iraq, which involved several recently-retired generals and admirals sharply criticizing Secretary of Defense Rumsfeld's leadership dynamics [@cook_revolt_2008].

My personal motivation in choosing the intended topic, however, stems from what I believe are changes in the type of civil-military disagreement that have surfaced in recent years. For example, I believe we are seeing portions of the military institution (particular retired military officers) openly disagree with civilian leaders on topics that I believe are, even by the widest of academic definitions, way outside of appropriate military expertise.^[I hope to prove this, empirically, in or as part of a chapter in my dissertation] More interestingly, these conflicts are occurring not in the context of major war, but the opposite: a period in which the levels of US Troops deployed in active combat zones is less than it has been in recent years. I further believe that while such developments are catching the attention of civil-military relations scholars, most of the work done on this topic is impressionistic [@golby_jim_keep_2020] as opposed to systematic. In other words, civil-military relations scholars have identified that polarized partisan politics serve as a risk for the military, but none to my knowledge have have specifically set out to systematically explore if and how polarized partisan politics are impacting the military's professional identity and character.

My intended dissertation will help answer several related questions, including: if personal and partisan loyalties are changing as a result of political polarization [need cite], is this having an impact within the military, i.e. is the concept of organizational loyalty changing as a result of polarization? Are these changes in organizational loyalty having an impact on the military's professional identity and character? Through what mechanisms do these changes operate? What sorts of changes are they driving, if any, in the military institution? in the military changing as a result of polarized domestic politics? If so, through what mechanisms, and how much? Is this problematic for the military profession, and why? The answers to these questions, I believe, will help inform the broader field of international relations on the broader and much more holistic issue of how changes in political polarization impact the military's ability to defend the nation and achieve the goals of American foreign policy.

# The Analytical Framework
## Polarization and Military Professionalism
